---
title: "Project title"
subtitle: "Proposal"
format: html
editor: visual
---

```{r}
#| label: load-pkgs
#| message: false

library(tidyverse)
```

# Data 1

## Introduction and data

-   Identify the source of the data.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

-   Write a brief description of the observations.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

-   Identify the types of variables in your research question. Categorical? Quantitative?

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-1
AAAW_national_covid_history <- read_csv("data/AAAW_national-covid-history.csv")

```

# Data 2

## Introduction and data

-   Identify the source of the data.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

-   Write a brief description of the observations.

## Research question

-   Research question: How does squad depth effect the play-off results of the NBA championship?

-   Importance: Depth is an often overlooked variable that significantly contributes to the over-all strength of the team. Performing an analysis on this question allow us to develop a systematic method to better understand and assess this often-neglected factor.

-   Description & hypotheses: This research aims to develop a systematic way to evaluate team depth using the NBA play-off data from 1950-2022. We will then investigate its effect on the final results of the championship. To evaluate team depth, we will look at variables such as the points by bench players, numbers of bench players, field goal percentage of role players. We will also ask questions such as are the player minutes evenly-spreaded, and are all the positions well-subsituted, to evaluate the depth of the team. Eventually, we will assign relative scores for team depth and evaluate its association with final championship results. We hypothesize that the deeper the team is, the more likely it is going to get higher places because it has plenty of substitutes to get players rest.

-   Identify the types of variables: We will use a mix of categorical and numerical variable to answer this questions. We will look into categorical variables such as player positions and championship results to evaluate the completeness and sucess of the team. We will look into numerical variables such as player minutes, field goal percentage, and numbers of bench player to eventually create a new numerical score that describes the team's overall depth.

## Literature

-   Published credible article: "Predicting the Outcome of NBA Playoffs Based on the Maximum Entropy Principle" (Cheng et al.)

-   Cheng et al. developed a machine learning model that utilized maximum entropy principle to predict the outcome of nba play-off results using 14 basic technical features, such as 3 points, field goals, and number of steals. Their model reached an accuracy of 74.4%, which outperformed the previous state-of-the-art with 70% accuracy. The author reasoned that this increase of performance came from the maximum entropy principle as well as their consideration of the team's last 6 game's data which gave a more detailed depiction of team's strength at the specific time before the play-off game.

-   Among the 14 features that they considered, team depth was not included. We believe that the team depth is a great factor that impacts the result of the play-off games. And if our hypothesis is true, the team depth variable could be a valuable addition to Cheng et al.'s machine learning to improve their accuracy.

## Glimpse of data

```{r}
#| label: load-data-2

AAAW_playoff_stats <- read_csv("data/AAAW_updatedplayoffStats.csv")
glimpse(AAAW_playoff_stats)
```

# Data 3

## Introduction and data

-   Identify the source of the data.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

-   Write a brief description of the observations.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

-   Identify the types of variables in your research question. Categorical? Quantitative?

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-3
AAAW_covid_cause <- read_csv('data/Conditions_Contributing_to_COVID-19_Deaths__by_State_and_Age__Provisional_2020-2023.csv')
```
