---
title: "Project title"
subtitle: "Proposal"
format: html
editor: visual
---

```{r}
#| label: load-pkgs
#| message: false

library(tidyverse)
```

# Data 1

## Introduction and data

-   Identify the source of the data

    The COVID Tracking Project is collecting this data.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

    This data is taken from the websites of state/territory public health authorities. The website states that when there are holes in the data, they use information from official state social media accounts and official announcements and statements by governors or other authorities. Some state's public health dept pages had covid specific pages that had this information.

-   Write a brief description of the observations.

    As time progressed throughout the pandemic, many more tests began being administered. This is evident in the new tests column as the amount of new tests  has a general upward trend. The correlation between new cases (positive tests) and negative PCR tests seems relatively consistent over time. It seems that there are usually 1/3-1/2 the amount of positive tests as there are negative tests. I see that there was a spike in hospitalization in December of 2020. 

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

    Is there a significant correlation between new tests administered on a certain day and deaths on that same day and each day up to two weeks after?

    On what day was there the highest amount of positive tests? And in the four weeks following, was the number of deaths greater than average?

    -   The first question is important because there were a lot of controversies raised during the pandemic about testings correlation with death rates. Many believed that mass testing was an effective way to reduce death rates. Many believed that testing was unnecessary and was not a beneficial use of resources. This question would answer if the amount of tests administered impact the amount of COVID deaths and if increased testing reduces the death rate over a span of the two week contagion period. 

    -   The second question responds to a similar controversy. Many were arguing that positive cases did not correlate with death rates and that there shouldn't be work done to decrease the amount of positive tests because it wouldn't have a significant health impact on the population. This question would follow the death rate the highest positive test day for the four weeks following that date. This is the time span that most die of the illness after contracting it. If the people arguing against intervention are correct, the average deaths for those days following the greatest positive test day won't be greater than that of the rest of the data for all the other days. 

-   A description of the research topic along with a concise statement of your hypotheses on this topic?

    The research topic that we are addressing with this data set is an analysis of common COVID controversies. We are attempting to discern if death rates had any correlation with testing rates. 

-   Identify the types of variables in your research question. Categorical? Quantitative?

    -   Categorical: There is no categorical data for this set 

    -   Quantitative: number of deaths a day, number of tests administered a day, number of positive tests, number of negative tests, hospitalization numbers 

## Literature

-   Find one published credible article on the topic you are interested in researching.

    <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7302472/pdf/mm6924e2.pdf>

-   Provide a one paragraph summary about the article.

This case surveillance report tracked all US COVID cases from January 22-May 30th of 2020. This report dives more into the demographics of those testing positive for COVID vs the nature of our data set which just gives the exact numerical reports for amount of deaths/tests/etc. This article specifically focuses on underlying conditions that COVID patients had and how that correspond to the severity of their illness (hospitalization vs intensive care vs those who died). It explains the median age of those contracting COVID. It also explains who was most likely to experience the illness asymptomatically. It provides data tables to discern the exact amount of people with pre-existing conditions and sorts them by sex and age. There is a data table reporting cumulative incidence by sex and age. There is also a graph that presents the number of cases and another one with the number of deaths during that same time frame. 

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

    This article, although covering more specificities, does not present monthly testing numbers throughout the peak of the pandemic. This report explains the demographics and the underlying conditions of the patients, but does not give any timeline for the AMOUNT of people testing positive. It also gives no information about the amount of people getting tested or amount of those who test negative. It only explains the demographics of those who test positive and then what happens to them post positive test. The changing of testing, hospitalization, and death numbers is what our data set adds on to the conversation. 

## Glimpse of data

```{r}
#| label: load-data-1
AAAW_national_covid_history <- read_csv("data/AAAW_national-covid-history.csv")

```

# Data 2

## Introduction and data

-   Identify the source of the data.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

-   Write a brief description of the observations.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

-   Identify the types of variables in your research question. Categorical? Quantitative?

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-2

AAW_playoff_stats <- read_csv("data/AAAW_updatedplayoffStats.csv") 
```

# Data 3

## Introduction and data

-   Identify the source of the data.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

-   Write a brief description of the observations.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

-   Identify the types of variables in your research question. Categorical? Quantitative?

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-3
AAAW_covid_cause <- read_csv('data/Conditions_Contributing_to_COVID-19_Deaths__by_State_and_Age__Provisional_2020-2023.csv')
```
