---
title: "Project title"
subtitle: "Proposal"
format: html
editor: visual
---

```{r}
#| label: load-pkgs
#| message: false

library(tidyverse)
```

# Data 1

## Introduction and data

-   Identify the source of the data.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

-   Write a brief description of the observations.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

-   Identify the types of variables in your research question. Categorical? Quantitative?

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-1
AAAW_national_covid_history <- read_csv("data/AAAW_national-covid-history.csv")

```

# Data 2

## Introduction and data

-   Identify the source of the data.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

-   Write a brief description of the observations.

## Research question

-   Research question: How does squad depth effect the play-off results of the NBA championship?

-   Importance: Depth is an often overlooked variable that significantly contributes to the over-all strength of the team. Performing an analysis on this question allow us to develop a systematic method to better understand and assess this often-neglected factor.

-   Description & hypotheses: This research aims to develop a systematic way to evaluate team depth using the NBA play-off data from 1950-2022. We will then investigate its effect on the final results of the championship. To evaluate team depth, we will look at variables such as the points by bench players, numbers of bench players, field goal percentage of role players. We will also ask questions such as are the player minutes evenly-spreaded, and are all the positions well-subsituted, to evaluate the depth of the team. Eventually, we will assign relative scores for team depth and evaluate its association with final championship results. We hypothesize that the deeper the team is, the more likely it is going to get higher places because it has plenty of substitutes to get players rest.

-   Identify the types of variables: We will use a mix of categorical and numerical variable to answer this questions. We will look into categorical variables such as player positions and championship results to evaluate the completeness and sucess of the team. We will look into numerical variables such as player minutes, field goal percentage, and numbers of bench player to eventually create a new numerical score that describes the team's overall depth.

## Literature

-   Published credible article: "Predicting the Outcome of NBA Playoffs Based on the Maximum Entropy Principle" (Cheng et al.)

-   Cheng et al. developed a machine learning model that utilized maximum entropy principle to predict the outcome of nba play-off results using 14 basic technical features, such as 3 points, field goals, and number of steals. Their model reached an accuracy of 74.4%, which outperformed the previous state-of-the-art with 70% accuracy. The author reasoned that this increase of performance came from the maximum entropy principle as well as their consideration of the team's last 6 game's data which gave a more detailed depiction of team's strength at the specific time before the play-off game.

-   Among the 14 features that they considered, team depth was not included. We believe that the team depth is a great factor that impacts the result of the play-off games. And if our hypothesis is true, the team depth variable could be a valuable addition to Cheng et al.'s machine learning to improve their accuracy.

## Glimpse of data

```{r}
#| label: load-data-2

AAAW_playoff_stats <- read_csv("data/AAAW_updatedplayoffStats.csv")
glimpse(AAAW_playoff_stats)
```

# Data 3

## Introduction and data

-   National Center For Health Statistics (NCHS)

-   This data was last updated March 8, 2023 and created May 8, 2020. It is not specified how the data was collected, but it was likely by surveying COVID-19 infection and death reports that were given to the NCHS.

-   Each observation is a time start and end date and COVID-19 deaths classified by age group, condition, jurisdiction, year, and month.

## Research question

-   Question: What 3 groups of co-morbid conditions are most associated with mortality due to COVID-19 and how do these compare to deaths from COVID-19 without mention of other co-morbid conditions?

-   Importance: Identification of diseases and conditions that are most directly related to mortality from COVID-19 provides valuable information regarding how patients should be treated and which individuals should take extra precautions to avoid infection. Treatment plans can be catered to these conditions: for example, an individual that is suffering from a "dangerous" co-morbidity in the context of COVID-19 can be immediately or prophylacticly prescribed an antiviral like Nirmatrelvir/ritonavir (Pavlovid TM) to avoid complications and mortality if infected.

-   Our research topic will focus on determining the 3 co-morbid condition groups that are most related to COVID-19 mortality. We will analyze how condition groups impact the incidence of COVID-19 mortality. We hypothesize that respiratory illnesses will be the predominant co-morbidity because COVID-19 is a respiratory illness, sepsis will be second because it is a common complication of infection, and Vascular and Unspecified Dementia will be third because it is more common among older age groups that are more vulnerable to COVID-19 mortality.

-   COVID-19 deaths is quantitative while Condition Groups are Categorical.

## Literature

-   https://jamanetwork.com/journals/jama/fullarticle/2765184 . Richardson S, Hirsch JS, Narasimhan M, et al. Presenting Characteristics, Comorbidities, and Outcomes Among 5700 Patients Hospitalized With COVID-19 in the New York City Area. *JAMA.* 2020;323(20):2052--2059. doi:10.1001/jama.2020.6775

-   This study analyzes the risk of hospitalization by co-morbidity for 5,700 individuals in New York City from March 1, 2020 to April 1,2020. It has been shown that the majority of individuals that are ultimately hospitalized for COVID-19 are of older age and have various co-morbid conditions. The study found that the most common co-morbid conditions among individuals hospitalized with COVID-19 are Hypertension, Obesity, and Diabetes.

-   Our research question builds on this research to generalize the study to the U.S as a whole and determine risk of death rather than hospitalization alone.

## Glimpse of data

```{r}
#| label: load-data-3
AAAW_covid_cause <- read_csv('data/Conditions_Contributing_to_COVID-19_Deaths__by_State_and_Age__Provisional_2020-2023.csv')
```
